R Tools for ObsPack, Receptors and Footprints (rtorf) for processing atmospheric observations in NOAA-GML

### 2025-05-20 paper/paper.Rmd
adding some legacy code
setup

```{r}
datasetid <- "aircraft-insitu"
df <- obs_read_nc(index = index,
                  categories = datasetid,
                  att = FALSE,
                  solar_time = FALSE,
                  verbose = TRUE)
```

@@

Now we have a `data.table` that contains `r ncol(df)` columns and 
`r nrow(df)` observations. The size of `data.table` is 
`r format(object.size(df), unit = "Gb")`. We can add a column
of time in format "POSIXct" and cut the seconds every 20 seconds.
Usually, aircraft observations every 1 second. Then,
We can simplify the data by calculating averages every 20 seconds.
We perform this task by cutting time very 20 seconds. Then,
we add a new column with the mandatory name `key_time`, which 
will be used to aggregate data with "POSIXct" class, but every 20 seconds.

@@

Now we have the data processed and ready to be exported. `rtorf` includes 
a number of functions to save the data as takbulated format in text, csv and 
CSVY[^1] are csv files with a YAML header. This funcitons can be seen in the documentation.
In this last part of the manuscript we will show some visualizations. We included
a function named `obs_plot` which plots data in long format using R base functions.
Here we see data for the month of March 2020. This usefull function allows to plot several
sites and prints the x-axis range.

@@

```{r}
x <- df3
x$ch4 <- x$value*1e+9
obs_plot(x, 
         time = "ch4", 
         y = "altitude_final", 
         colu = "month", 
         type = "b", 
         xlab = expression(CH[4]~ppb), 
         ylab = "altitude (m)")
```

>>>>

```{r read_air}
datasetid <- "aircraft-insitu"
df <- obs_read_nc(index = index,
                  categories = datasetid,
                  att = FALSE,
                  solar_time = FALSE,
                  verbose = TRUE)
```

@@

Now we have a `data.table` that contains `r ncol(df)` columns and 
`r nrow(df)` observations. The size of `data.table` is 
`r format(object.size(df), unit = "Gb")`. Sometimes the data can be alaready
filtered every 20 seconds or for a different period of time.
However, raw data can be availalb ein a second-by-second basis. 
Under this case we may need to aggregate datah. In this example, we can add a column
of time in format "POSIXct" and cut the seconds every 20 seconds.
Usually, aircraft observations every 1 second. Then,
We can simplify the data by calculating averages every 20 seconds.
We perform this task by cutting time very 20 seconds. Then,
we add a new column with the mandatory name `key_time`, which 
will be used to aggregate data with "POSIXct" class, but every 20 seconds.

@@

Now we have the data processed and ready to be exported. `rtorf` includes 
a number of functions to save the data as tabulated format in text, csv and 
CSVY[^1] are csv files with a YAML header. This funcitons can be seen in the documentation.
In this last part of the manuscript we will show some visualizations. We included
a function named `obs_plot` which plots data in long format using R base functions.
Here we see data for the month of March 2020. This usefull function allows to plot several
sites and prints the x-axis range.

@@

```{r}
x <- df3
x$ch4 <- x$value*1e+9
obs_plot(x, 
         time = "ch4", 
         y = "altitude_final", 
         colu = "month", 
         type = "b", 
         xlab = expression(CH[4]~ppb), 
         ylab = "altitude (m)")
```
# HYSPLIT
